BIOL 274 Homework 2

### 2024-01-31 bio_dynamics/bio274_hw2.Rmd
bio274 hw2

---
title: "BIOL 274 Homework 2"
author: "Ruby Krasnow"
date: "2024-01-27"
output: pdf_document
---

@@

Although since we are working in discrete time, it might be more accurate to write the change in population as a difference equation, $P_{n+1}-P_n=\Delta P= 0.3P \left( 1- \frac{P}{2500} \right)$. For now we will continue using differential equation notation and apologize to Leibniz.

The initial conditions will remain the same for each of the three situations, but we can adjust the differential equation for each scenario as follows:

@@

Now we're going to have some fun and see how the population would change over a 100-year period if no harvesting occurred and under each of these three harvesting strategies! We will use the numerical ODE solver from the `deSolve` package.

@@

Let's see what we got!
```{r}
ggplot()+
  geom_line(data=fish_output, aes(x=time, y=P, color=names_full))+
  theme_bw()+
  labs(x="Time (years)", y="Pop. size", color=NULL)
```

@@

Our final solution is therefore:
\begin{equation}\label{solution}P(t)=\frac{2500}{10e^{\frac{t}{30}}-9}\end{equation}
Here is what it looks like:
```{r}
t=c(1:250)
y=2500/(10*exp(t/30)-9)

@@

```{r}
dy=D(expression(2500/(10*exp(t/30)-9)), "t")

@@

```{r}
N0 = 0.1
a=0.003
p0_small=10
p0_big=50

@@

## Question 4 & 5
For problems 4 and 5, use Euler’s Method with the given step size, h, to approximate the solution
to the given IVP over the time interval specified. Your answer should include a table of
approximate values of the dependent variable. It should also include a sketch of the graph
of the approximate solution.

>>>>

---
title: "BIOL 274 Homework 2"
author: "Ruby Krasnow"
date: "2024-01-31"
output:
  pdf_document:
    extra_dependencies: ["derivative", "tcolorbox", "xcolor"]
---

@@

Note: Since we are working in discrete time, it might be more accurate to write the change in population as a difference equation, $P_{n+1}-P_n=\Delta P= 0.3P \left( 1- \frac{P}{2500} \right)$. For now we will continue using differential equation notation and apologize to Leibniz.

The initial conditions $P(0)=2500$ will remain the same for each of the three situations, but we can adjust the differential equation for each scenario as follows:

@@

Let's see how the population would change over a 100-year period if no harvesting occurred and under each of these three harvesting strategies! We will use the numerical ODE solver from the `deSolve` package.

@@

Let's see what we got!
```{r,   fig.height=3, fig.width=5, fig.width=5}
ggplot()+
  geom_line(data=fish_output, aes(x=time, y=P, color=names_full))+
  theme_bw()+
  labs(x="Time (years)", y="Pop. size", color=NULL)
```

@@

Our final solution is therefore:
\begin{equation}\label{solution}P(t)=\frac{2500}{10e^{\frac{t}{30}}-9}\end{equation}
Here is what it looks like:
```{r,   fig.height=3, fig.width=5  }
t=c(1:250)
y=2500/(10*exp(t/30)-9)

@@

```{r,   fig.height=3, fig.width=5  }
dy=D(expression(2500/(10*exp(t/30)-9)), "t")

@@

```{r,   fig.height=3, fig.width=5  }
N0 = 0.1
a=0.003
p0_small=10
p0_big=50

@@

## Euler's Method
For problems 4 and 5, use Euler’s Method with the given step size, h, to approximate the solution
to the given IVP over the time interval specified. Your answer should include a table of
approximate values of the dependent variable. It should also include a sketch of the graph
of the approximate solution.

### Question 4
$$\frac{dw}{dt}=(3-w)(w+1), \quad w(0)=4, \quad 0\leq t \leq 5, \quad h=1.0$$
We start with a table that looks like this:
```{r}
data.frame(t=c(0:5), w=c(4, rep(NA, 5)))
```

**Manual computation**  
At $t=1, \quad w=4+(3-4)(4+1)1 = 4+(-5)=-1$  
But then for $t=2, \quad w=-1+(3--1)(-1+1)1= -1+0 = -1$  
For $t=3, w=-1+(3--1)(-1+1)1= -1+0 = -1$  
So once we hit $t=2$, the slope is always 0, making $y(t_n+1) = y(t_n) = -1$. The same thing will happen each time we compute $w$, so using Euler's method with a step size of h=1 has gotten us stuck!

**Using a function**  
Let's define a function to compute Euler's method and see if we can do better with smaller step sizes.

Here we will write our own function, but note that the `deSolve` package includes a function `euler` that can do the same thing with many options for additional customization. There is also a simpler `euler` function in the package `demodelr` that takes fewer parameters.

```{r}
euler_fun <- function(slope_fun, x0, y0, h, N) {
    x <- x0
    y <- y0
    num_steps <- N/h

    for(i in 1:num_steps) {
        y0 <- y0 + h * slope_fun(x0, y0)
        x0 <- x0 + h
        x <- c(x, x0)
        y <- c(y, y0)
    }

    return(data.frame(x = x, y = y, h=h))
}

f <- function(x_n, y_n){
  slope=(3-y_n)*(y_n+1)
  slope
}

euler_fun(slope_fun=f, x0=0, y0=4, h=1, N=5) %>% rename(w=y, t=x)
euler_fun(slope_fun=f, x0=0, y0=4, h=0.5, N=5) %>% rename(w=y, t=x)
euler_fun(slope_fun=f, x0=0, y0=4, h=0.1, N=5)%>% rename(w=y, t=x)

euler_output<- c(1, 0.5, 0.1) %>% 
  map(\(x) euler_fun(f, x0=0, y0=4, h=x, N=5)) %>% 
  bind_rows()

```

```{r,   fig.height=3, fig.width=5  }
ggplot(data = euler_output)+
  geom_point(aes(x=x, y=y, color=as.factor(h)))+
  geom_line(aes(x=x, y=y, color=as.factor(h)), alpha=0.5)+
  theme_light()+
  labs(x="t", y="w", color="h")
```

Since this is a separable ODE and we are given initial conditions, we can also find the particular solution explicitly:
$$\frac{dw}{dt}=(3-w)(w+1)=-w^2+2w+3$$

$$\int \frac{dw}{w^2-2w-3}=\int -dt$$

Using partial fraction decomposition for the left-hand side, we can rewrite the integral as:
$$\frac{1}{4}\int \frac{1}{w-3}-\frac{1}{w+1}\>dw=\int -dt$$

$$\frac{1}{4}\left( \ln(w-3)-\ln(w+1) \right)=-t+C_1$$
$$ \ln\left(\frac{w-3}{w+1}\right) =-4t+C_2$$

$$\frac{w-3}{w+1}=\frac{w+1-4}{w+1}=1+\frac{-4}{w+1}=C_3e^{-4t}$$
$$\frac{-4}{w+1}=C_3e^{-4t}-1$$
Plugging in our initial conditions, 
$$\frac{-4}{5}=C_3-1 \quad \longrightarrow \quad C_3=\frac{1}{5}$$

$$\frac{-4}{w+1}=\frac{1}{5}e^{-4t}-1$$
\begin{tcolorbox}[colback=white, title=Solution]
$$w(t)=\frac{4}{1-\frac{1}{5}e^{-4t}}-1$$
\end{tcolorbox}
\begin{tcolorbox}[colback=blue!5!white,colframe=blue!75!black,title=Check solution by differentiating]

Direct differentiation of solution
$$\frac{dw}{dt}= -4\left ( 1-\frac{1}{5}e^{-4t} \right)^{-2}\left(\frac{-1}{5}e^{-4t}\right)(-4) = \frac{16 \left(- \frac{1}{5} e^{-4t} \right)}{ \left(1 -\frac{1}{5}e^{-4t} \right)^2}$$
Plugging solution into original ODE

$$\frac{dw}{dt}=(3-w)(w+1)=\left(3-\frac{4}{1-\frac{1}{5}e^{-4t}}+1 \right) \left(\frac{4}{1-\frac{1}{5}e^{-4t}}-1+1 \right) =$$

$$\left(4-\frac{4}{1-\frac{1}{5}e^{-4t}}\right)\left(\frac{4}{1-\frac{1}{5}e^{-4t}}\right) = $$

$$\frac{16}{1-\frac{1}{5}e^{-4t}}-\frac{16}{\left(1-\frac{1}{5}e^{-4t}\right)^2} =
\frac{16 \left( 1 - \frac{1}{5} e^{-4t} \right) -16}{ \left(1 -\frac{1}{5}e^{-4t} \right)^2}=$$
$$\frac{16 \left(- \frac{1}{5} e^{-4t} \right)}{ \left(1 -\frac{1}{5}e^{-4t} \right)^2}$$

\end{tcolorbox}

So now we can add to our graph the actual solution:
```{r}
actual_eqn <- data.frame(t=seq(from=0, to=5, by=0.0001)) %>% mutate(
w=4/((1-(1/5)*exp(-4*t)))-1)

ggplot()+
  geom_line(data=actual_eqn, aes(x=t, y=w, color="Actual solution"), linewidth=1.5)+
  geom_point(data = euler_output, aes(x=x, y=y, color=as.factor(h)), size=1)+
  geom_line(data = euler_output, aes(x=x, y=y, color=as.factor(h)), alpha=0.5, linewidth=1)+
  theme_light()+
  labs(x="t", y="w", color="h")+
  scale_color_manual(values=c( "#e69b99","#89689d","#015b58", "black"))
#Credit: hex codes are from Jake Lawlor's R package PNWcolors
```

We can also see that $\lim\limits_{t \to \infty}w(t)=3$ by looking at our expression for the actual solution, because $e^{-4t}$ will approach zero, so the denominator will approach $1-0$ and the overall solution will approach $\frac{4}{1}-1=3$. Clearly, the step size of $h=0.1$ very closely approximated the actual solution after around 30 time steps.

### Question 5
$$\frac{dy}{dt}=e^{2/y}, \quad y(0)=2, \quad 0\leq t \leq 2, \quad h=0.5$$

**Manual computation**

$(t_0=0, y_0=2) \quad \longrightarrow \quad$ slope $=e^{2/2}=e, y_1=2+\frac{e}{2} \approx$ `r exp(2+exp(1)/2)`
```{r}
y_1=round(2+exp(1)/2, 4)
y_1
```

$(t_1=0.5, y_1 \approx 3.3591)\quad \longrightarrow \quad$ slope $=\exp{\left(\frac{2}{3.3591}\right)} \approx$ `r exp(2/3.3591)`
```{r}
y_2=round(3.3591+exp(2/3.3591)/2, 4)
y_2
```

$(t_2=1, y_2 \approx 4.266)\quad \longrightarrow \quad$ slope $=\exp{\left(\frac{2}{4.266}\right)} \approx$ `r exp(2/4.266)`
```{r}
y_3=round(4.266+exp(2/4.266)/2, 4)
y_3
```

$(t_3=1.5, y_3 \approx 5.0651)\quad \longrightarrow \quad$ slope $=\exp{\left(\frac{2}{5.0651}\right)} \approx$ `r exp(2/5.0651)`

```{r}
y_4=round(5.0651+exp(2/5.0651)/2, 4)
y_4
```

$(t_4=2, y_4\approx 5.8072)$

Using our function from earlier,
```{r}
new_f = function(x_n, y_n) {
  slope = exp(2/y_n)
  slope
}

euler_fun(slope_fun = new_f, h=0.5, N=2, x0=0, y0=2)
```

And we get the same result as our manual computation.

```{r}
library(PNWColors)

euler_output_2<- c(1, 0.5, 0.1, 0.05) %>% 
  map(\(x) euler_fun(new_f, x0=0, y0=2, h=x, N=2)) %>% 
  bind_rows()

ggplot()+
  geom_point(data = euler_output_2, aes(x=x, y=y, color=as.factor(h)))+
  geom_line(data = euler_output_2, aes(x=x, y=y, color=as.factor(h)), alpha=0.8)+
  theme_light()+
  labs(x="t", y="y", color="h")+
  scale_color_manual(values=pnw_palette(name="Starfish",n=4,type="discrete"))
```
